Bayesian statistics and swine flu

### 2009-08-13 docs/pandemic/bayesPandemic.rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@359 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<setup,fig=false>>=
options(SweaveHooks=list(fig=function() par(mar=c(2.5,2.5,0.1,0.1), mgp=c(1.5, 0.5, 0), cex=0.6)))
# source in all the R files
library(pandemic)
@

@@

\begin{frame}
\frametitle{The Problem}
\begin{itemize}
	\item Some people will get infected with swine flu
	\item Some will get hospitalized, some will die
	\item How do we fatality rate and the hospitalization rate?
	\item Problem: censoring, some individuals in the dataset have not yet had their hospitalization and/or fatality
	\item Crude rates underestimate the rates, and result in estimates increasing over time
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Types of infection}

@@

\begin{frame}
\frametitle{Distributions}
\begin{itemize}
	\item $f(t;\mu,\nu,\delta)$ is the distribution for time to an event.  
	\item \ldots a rounded, zero-inflated Weibull
\begin{align*}
Z\sim&\text{Bern}(\delta)\\
[Y|Z=1] =& 0 \text{ w.p.} 1\\
pr(T=t|Z=0) =&  \text{pr}(\max(t-0.5,0) < y < t+0.5)\\
y \sim & \text{Weibull}(\mu/\Gamma(1 + 1/\nu), \nu)
\end{align*}
\item $\delta$ is the probability of transitioning immediately for an administrative (not biological) reason, 
\begin{itemize}
	\item i.e.\ arriving dead at the hospital.
\end{itemize}
\item Barring administrative zeros, $\mu = E(T|Z=0)$ is the average time to an event
\begin{itemize}
	\item $\mu_I$ is average time to infection
\end{itemize}
\item $\nu$ is a Weibull shape parameter
\begin{itemize}
	\item $\nu=1$ is constant hazards
	\item $\nu > 1$ results in more regular event times.
\end{itemize}

@@

\begin{frame}
\frametitle{Priors}
\begin{itemize}
	\item What does expert opinion tell us these parameters should be?
	\item Based on previous experience, not the current data.
	\item We need a distribution, not a single value
\begin{itemize}
	\item If we `knew' the correct value, there would be no need for this model
\end{itemize}
\item Gamma or Beta priors, parametrized by mean and standard deviation
\item Average time to onset of symptoms is expected to be 3.5 days, but the average time could be a day longer or shorter.
\begin{itemize}
	\item $E(\mu_I = 3.5)$ and $sd(\mu_I = 0.5$
	\item $\mu_I \sim \text{Gamma}(3.5^2/0.5^2 , 3.5/0.5^2 )$
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
	\item $\rho_D$ is the fatality ratio, or probability of a deadly infection
	\item $\rho_S$ is the proportion of non-fatalities which are hospitalized
	\item so the probability of being hospitalized is $\rho_D + (1-\rho_D)\rho_S$
	\item parametrize this way to ensure the probabilities of being Mild, Severe, or Deadly add to 1.
	\item beta priors on $\rho_D$ and $\rho_S$, parametrized by mean and standard deviation
	\item i.e.\ $\rho_D$ is expected to be $0.1$, with $sd(\rho_D = 0.05)$
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{What we need}
\begin{alltt}
\input{bayesPandemicPriors.txt}
\end{alltt}
\end{frame}

@@

\begin{frame}
\frametitle{The Algorithm}
data augmentation

random walk metropolis on parameters
\end{frame}

\begin{frame}[fragile]
\frametitle{Simulated data}
<<simData,echo=true,fig=false>>=
params = pandemicParams( 
  MedHospD = c(mean = 1.5, shape = 2, zeros = 0.2)
  )

data = simEpidemic(params, delta=3, days=20)

data[1:10,-1]
@

@@

\begin{frame}[fragile]
\frametitle{Results}
<<mcmcResults,echo=true,fig=false>>=
paramSample = mcmcPandemic(data,params,priors,sigma=0.1,runs=100)
dim(paramSample)
colnames(paramSample)
@
\end{frame}

@@

\begin{block}{Histogram}
<<histPost>>=
hist(paramSample[,"probs.D"], breaks=12)
abline(v=params$probs["D"], col="red")
@
\end{block}
\end{columns}
\end{frame}
\end{document}

\begin{frame}
\frametitle{Age?}

piecewise constant 5yr?

change points?

note that after data augmentation it's a standard problem.

@@

\begin{frame}
\frametitle{To do}
\begin{itemize}
	\item simulate hospital loads
	\item model the infection process
	\item spatial infection process
	\item covariates

\end{itemize}
\end{frame}

@@

>>>>

<<setup,fig=false>>=
options(SweaveHooks=list(fig=function() par(mar=c(2.5,2.5,0.1,0.1), mgp=c(1.5, 0.5, 0), cex=0.6)), width=58)
# source in all the R files
library(pandemic)
@

@@

\begin{frame}
\frametitle{The Problem}
\begin{itemize}
	\item Some people will get infected with swine flu
	\item Some will get hospitalized, some will die
	\item How do we estimate the fatality and hospitalization rates?
	\item Problem: censoring, some individuals in the dataset have not yet had their hospitalization and/or fatality
	\item Crude rates underestimate the rates, and result in estimates increasing over time
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Types of infection}
All infections are one of the following types:
\begin{description}
	\item[Mild:] The individual will recover without needing to be hospitalized
	\item[Serious:] The infection is severe enough that hospitalization will be required before the individual recovers
	\item[Deadly:] Following hospitalization, the patient dies of the infection.
\end{description}
When a case is first identified, it's not known what type it is.

@@

\begin{frame}
\frametitle{Distributions}
\begin{itemize}
	\item $ f(t;\mu,\nu,\delta) = pr(T=t)$ is the distribution for time to an event.  
	\item \ldots a rounded, zero-inflated Weibull
\begin{align*}
Z\sim&\text{Bern}(\delta)\\
[T|Z=1] =& 0 \text{ w.p.} 1\\
pr(T=t|Z=0) =&  \text{pr}(\max(t-0.5,0) < y < t+0.5)\\
y \sim & \text{Weibull}(\mu/\Gamma(1 + 1/\nu), \nu)
\end{align*}
\item $\delta$ is the probability of transitioning immediately for an administrative (not biological) reason, 
\begin{itemize}
	\item i.e.\ arriving dead at the hospital.
\end{itemize}
\item Barring administrative zeros, $\mu = E(T|Z=0)$ is the average time to an event
\begin{itemize}
	\item $\mu_I$ is average time to infection
\end{itemize}
\item $\nu$ is a Weibull shape parameter
\begin{itemize}
	\item $\nu=1$ is constant hazards
	\item $\nu > 1$ results in more regular event times.
\end{itemize}

@@

\begin{frame}
\frametitle{Priors}
\begin{itemize}
	\item What does expert opinion tell us these parameters should be?
	\item Based on previous experience, not the current data.
	\item We need a distribution, not a single value
\begin{itemize}
	\item If we `knew' the correct value, there would be no need for this model
\end{itemize}
\item Gamma or Beta priors, parametrized by mean and standard deviation
\item Average time to onset of symptoms is expected to be 3.5 days, but the average time could be a day longer or shorter.
\begin{itemize}
	\item $E(\mu_I = 3.5)$ and $sd(\mu_I)	 = 0.5$
	\item $\mu_I \sim \text{Gamma}(3.5^2/0.5^2 , 3.5/0.5^2 )$
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
	\item $\rho_D$ is the fatality ratio, or probability of a deadly infection
	\item $\rho_S$ is the proportion of non-fatalities which are hospitalized
	\item so the probability of being hospitalized is $\rho_D + (1-\rho_D)\rho_S$
	\item parametrize this way to ensure the probabilities of being Mild, Severe, or Deadly add to 1.
	\item beta priors on $\rho_D$ and $\rho_S$, parametrized by mean and standard deviation
	\item i.e.\ $\rho_D$ is expected to be $0.1$, with $sd(\rho_D) = 0.05$
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{What we need}
{\footnotesize
\begin{alltt}
\input{bayesPandemicPriors.txt}
\end{alltt}
}
\end{frame}

@@

\begin{frame}
\frametitle{The Algorithm}
\begin{block}{Data augmentation}
\begin{itemize}
	\item Fix the parameter values
	\item Simulate the missing data: infection type, times of infection, onset, hospitalization, recovery, and/or death.
	\item Can simulate directly, as conditioning is easy.
\end{itemize}
\end{block}

\begin{block}{Random walk metropolis}
\begin{itemize}
	\item With the full augmented data, the likelihood is simple as observations are independent.
	\item Propose new parameters with a Gaussian random walk, accept with the appropriate probability.
\end{itemize}
\end{block}

@@

\begin{frame}[fragile]
\frametitle{Simulated data}
<<simData,echo=true,fig=false>>=
params = pandemicParams( 
	InfOns = c(mean=1.5, shape=2, zeros=.1),
	OnsMedM= c(mean=1, shape=1, zeros=.1 ),
	OnsMedS= c(mean=1, shape=1, zeros=.1 ),
	OnsMedD= c(mean=1, shape=1, zeros=.1 ),
	MedRec = c(mean=3, shape=2, zeros=.1, lost=0.95 ),
  MedHospS = c(mean = 2, shape = 2, zeros = 0.1),
  MedHospD = c(mean = 2, shape = 2, zeros = 0.1),
	HospRec = c(mean=2, shape=1.5, zeros=.01 ),
	HospDeath = c(mean=2, shape=0.1, zeros=.05 ),
 	probs = c(M=0.7, S=0.25, D=0.05)
  )

data = simEpidemic(params, delta=4, days=20)

data[1:8,-1]
@

\begin{frame}[fragile]
\frametitle{Results}
<<mcmcResults,echo=true,fig=false>>=
paramSample = mcmcPandemic(data,params,priors,sigma=0.05,runs=600)
paramSample = paramSample[seq(100,dim(paramSample)[1], by=5),]
dim(paramSample)
colnames(paramSample)
@
\end{frame}

@@

\begin{block}{Histogram}
<<histPost>>=
hist(paramSample[,"probs.D"], breaks=12)
abline(v=params$probs["D"], col="red")
@
\end{block}
\end{columns}
\end{frame}

\begin{frame}[fragile]
\frametitle{Summary Statistics}
<<summaryStats,fig=false>>=
postSummary=t(apply(paramSample, 2, function(qq) 
  c(mean=mean(qq), quantile(qq, prob=c(0.025, 0.5, 0.975)))))
postSummary = cbind(true=unlist(params)[rownames(postSummary)], postSummary)

postSummary[1:12,-4]
@
\end{frame}

\begin{frame}
\frametitle{Posterior and prior, Hosp $\Rightarrow$ Death }
<<priorPost, fig=false,results=tex>>=
library(ellipse)
plotPrior(priors, paramSample, transition="MedHospS",file="MedHospS", tex=TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Hospital demand}
<<hospLoad,echo=T>>=
# generate hospital times
hospSample = simHospitals(paramSample , Ndays=20, NinfectionsPerDay=20)

# graph hospital times
plotHospitals(hospSample)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{With rising and falling incidence}
{\scrpitsize
<<hospLoad2,echo=T,height=2,width=4>>=
# generate hospital times
hospSample2 = simHospitals(paramSample , Ndays=40, 
	NinfectionsPerDay=c(seq(from=1,by=2, len=10), seq(from=20, len=10, by=2), rep(2, 20)))

# graph hospital times
plotHospitals(hospSample2)
@
}
\end{frame}

\begin{frame}
\frametitle{Age?}
\begin{itemize}
	\item Assume the fatality and hospitalization probabilities vary with age
	\item After data augmentation, we have a binary response with a smoothly-varying probability
	\item Use a canned Bayesian semi-parametric routine to generate a new age curve at every iteration.
	\item {\tt DPpackage}?
	\item Informative prior on the average or rate at a baseline age
	\item Strong prior that the curve is flat.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{To do}
\begin{itemize}
	\item more covariates	
	\item model the infection process
	\item spatial infection process
\end{itemize}
\end{frame}

\begin{frame}

\frametitle{Conclusions}

\begin{itemize}
	\item Use the data even if there isn't much of it
	\item If there are many cases and no fatalities, at least we can get an upper bound on the fatality rate
	\item When data are plentiful and informative, there is a case to be made for non-parametric survival methods
	\item But if rates change with age, sex, geography, or existing health conditions, even late stage epidemic data might not be sufficient for standard survival methods.
\end{itemize}

@@

@@

D75A0BD6A6A1
